---
title: "Climate Visualisations"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: united
    code_folding: hide
    toc: true
    toc_float: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(scales)
library(glue)
```

## GISS Surface Temperature Analysis (GISTEMP v4)

The GISS Surface Temperature Analysis ver. 4 (GISTEMP v4) is an estimate of global surface temperature change.

### Temperature Index Plot

```{r Temperature Index Plot}

read_csv("data/GLB.Ts+dSST.csv", skip = 1, na = "***") %>% 
  select(year = Year, t_diff = `J-D`) %>% 
  drop_na() %>% 
  ggplot(aes(x = year, y = t_diff)) +
  geom_line(aes(colour = "1"), size = 0.5, show.legend = FALSE) +
  geom_point(fill = "white", aes(colour = "1"), shape = 21, show.legend = TRUE) +
  geom_smooth(se = FALSE, aes(colour = "2"), size = 0.5, span = 0.2, show.legend = FALSE) + 
  scale_x_continuous(breaks = seq(1880, 2023, 20), expand = c(0,0)) +
  scale_y_continuous(limits = c(-0.5, 1.5), expand = c(0,0)) +
  scale_colour_manual(name = NULL, breaks = c(1,2), values = c("gray", "black"), labels = c("Annual mean", "Loess smoothing"), guide = guide_legend(override.aes = list(shape = 15, size = 5))) +
  labs(x = "", y = "Temperature Anomaly w.r.t 1951-1980 (\u00B0C)",
       title = "Global Annual Mean Surface Air Temperature Change",
       subtitle = "Data source: NASA's Goddard Institute for Space Studies (GISS).\nCredit: NASA/GISS") +
  theme_light() +
  theme(axis.ticks = element_blank(),
        axis.title.y = element_text(size = 9),
        plot.title.position = "plot",
        plot.title = element_text(margin = margin(b = 10), colour = "red", face = "bold"),
        plot.subtitle = element_text(size = 8, margin = margin(b = 10)),
        legend.position = c(0.15, 0.9),
        legend.title = element_text(size = 0),
        legend.key.height = unit(10, "pt"),
        legend.margin = margin(0,0,0,0))
```

GISTEMP Team, 2022: GISS Surface Temperature Analysis (GISTEMP), version 4. NASA Goddard Institute for Space Studies. Dataset accessed 2022-06-19 at https://data.giss.nasa.gov/gistemp/.

### Temperature Bar Plot

```{r Temperature Bar Plot}

t_data <- read_csv("data/GLB.Ts+dSST.csv", skip = 1, na ="***") %>%
  select(year = Year, t_diff = `J-D`) %>% 
  drop_na()

annotation <- t_data %>% 
  arrange(year) %>% 
  slice(1, n()) %>% 
  mutate(t_diff = 0,
         x = year + c(-5,5))

t_data %>% 
  ggplot(aes(x = year, y = t_diff, fill = t_diff)) +
  geom_col(show.legend = FALSE) +
  geom_text(data = annotation, aes(x = x, label = year), colour = "white") +
  geom_text(x = 1880, y = 1, hjust = 0,
            label = glue("Global temperatures have increased by over {max(t_data$t_diff)}\u00B0C since {min(t_data$year)}"),
            colour = "white") +
  scale_fill_stepsn(colours = c("darkblue", "white", "darkred"),
                    values = rescale(c(min(t_data$t_diff), 0, max(t_data$t_diff))),
                    limits = c(min(t_data$t_diff), max(t_data$t_diff)),
                    n.breaks = 9) +
  theme_void() +
  theme(plot.background = element_rect(fill = "black"))
```

### Temperature Stripe Plot

```{r Temperature Stripe Plot}

t_data <- read_csv("data/GLB.Ts+dSST.csv", skip = 1, na = "***") %>% 
  select(year = Year, t_diff = `J-D`) %>% 
  drop_na()

t_data %>% 
  ggplot(aes(x = year, y = 1, fill = t_diff)) +
  geom_tile(show.legend = FALSE) +
  scale_fill_stepsn(colours = c("#08306B", "white", "#67000D"),
                    values = rescale(c(min(t_data$t_diff), 0 , max(t_data$t_diff))),
                    n.breaks = 12) +
  coord_cartesian(expand = FALSE) +
  scale_x_continuous(breaks = seq(1890, 2020, 20)) +
  labs(title = glue("Global Temperature Change ({min(t_data$year)}-{max(t_data$year)})")) +
  theme_void() +
  theme(
    axis.text.x = element_text(colour = "white",
                               margin = margin(t = 5, b = 10, unit = "pt")),
    plot.title = element_text(colour = "white",
                              margin = margin(b = 5, t = 10, unit ="pt"),
                              hjust = 0.05),
    plot.background = element_rect(fill = "black"))

```

### Temperature Line Plot

```{r Temperature Line Plot}

t_diff <- read_csv("data/GLB.Ts+dSST.csv", skip = 1, na = "***") %>% 
  select(year = Year, month.abb) %>% 
  drop_na() %>% 
  pivot_longer(-year, names_to = "month", values_to = "t_diff") %>% 
  mutate(month = factor(month, levels = month.abb))

last_dec <- t_diff %>% 
  filter(month == "Dec") %>% 
  mutate(year = year + 1,
         month = "last_Dec")

next_jan <- t_diff %>% 
  filter(month == "Jan") %>% 
  mutate(year = year - 1,
         month = "next_Jan")

t_data <- bind_rows(last_dec, t_diff, next_jan) %>% 
  mutate(month = factor(month, levels = c("last_Dec", month.abb, "next_Jan")),
         month_number = as.numeric(month) - 1,
         this_year = year == 2022)

annotation <- t_data %>% 
  slice_max(year) %>% 
  slice_max(month_number)
  
t_data %>% 
  ggplot(aes(x = month_number, y = t_diff, group = year, colour = year, size = this_year)) +
  geom_hline(yintercept = 0, colour = "white") +
  geom_line() +
  geom_text(data = annotation,
            aes(x = month_number, y = t_diff, label = year, colour = year), 
            inherit.aes = FALSE,
            hjust = 0, size = 5, nudge_x = 0.15, fontface = "bold") +
  scale_x_continuous(breaks = 1:12,
                     labels = month.abb,
                     sec.axis = dup_axis(name = NULL, labels = NULL)) +
  scale_y_continuous(breaks = seq(-2, 2, 0.2),
                     sec.axis = dup_axis(name = NULL, labels = NULL)) +
  scale_size_manual(breaks = c(FALSE, TRUE),
                    values = c(0.25, 1), guide = "none") +
  scale_colour_viridis_c(breaks = seq(1880, 2020, 20),
                         guide = guide_colorbar(frame.colour = "white", 
                                                frame.linewidth = 1)) +
  coord_cartesian(xlim = c(1,12)) +
  labs(x = NULL,
       y = "Temperature change since pre-industrial times [\u00B0C]",
       title = glue("Global temperature change since {min(t_diff$year)} by month")) +
  theme(
    panel.background = element_rect(fill = "black", colour = "white", size = 1),
    plot.background = element_rect(fill = "#444444"),
    panel.grid = element_blank(),
    axis.text = element_text(colour = "white", size = 13),
    axis.ticks = element_line(colour = "white"),
    axis.ticks.length = unit(-5, "pt"),
    axis.title = element_text(colour = "white", size = 13),
    plot.title = element_text(colour = "white", size = 15, hjust = 0.5),
    legend.title = element_blank(),
    legend.background = element_rect(fill = NA),
    legend.text = element_text(colour = "white"),
    legend.key.height = unit(55,"pt")
  )

```

